---
title: "Two examples of using mbquartR"
author: "Alex Koiter"
date: "2025-02-04"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two examples of using mbquartR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)

options("mbquartR_example" = TRUE)
```


# Introduction

The  `mbquartR` package was created to help people easily identify, search, and locate parcels of land using the Manitoba legal descriptions. These legal land descriptions originate from the [Dominion Land Survey system](https://en.wikipedia.org/wiki/Dominion_Land_Survey), introduced in the late 19th century to organize the European settlement and colonization of Western Canada.

There are two main ways in which the `mbquartR` can be used.

1. Search for and locate parcels of land using legal land descriptions (e.g., SW-9-8-6E1)

or 

2. Search for legal land description using coordinates (latitude and longitude)

**Note:** The examples below are fictitious scenarios designed to highlight the functionality of the `mbquartR` package. The identified land parcels and coordinates are not intended for any official or practical use.

## Example 1: Using legal land description to find the locations of archived soil samples

Since 1926, the [Manitoba Soil Survey](https://www.gov.mb.ca/agriculture/soil/soil-survey/index.html) has been working to quantify and map soil properties (texture, parent material, stoniness, pH, and salinity etc) to support sustainable agricultural development across the province. Part of the soil survey has included collecting soil samples for chemical analysis (Carbon content). This has resulted in a rich archive of soil samples spanning more than eight decades. Recently there has been interest in reanalyzing these archived soil samples using new technology. However, the location of many older soil samples is identified solely by the quarter section description.

For many people, including myself, we are not familiar enough with land survey system to easily locate it. There are over 900,000 entries in the Manitoba Original Survey Legal Descriptions data covering more than 600,000 km<sup>2</sup>. Most mapping services are also unable to search for legal land description (e.g., Google Maps) which makes identifying where these soil sample were taken very difficult!

For an explanation of quarter sections and other land types, please visit the [Get started](https://docs.ropensci.org/mbquartR/articles/mbquartR.html) page.

### Scenario:

Here is a small data set that includes the soil organic carbon content (%) of four topsoil samples (0-15 cm depth) collected in 1980 as part of the Soil Survey at four different locations only identified by their legal land description. I would like to know where these soil samples were taken so I have some additional context for interpreting the data. Three of these samples were collected in Quarter Sections (most common type of land division) and one was collected in a River Lot

```{r}
soil_soc <- data.frame(
  soc = c(1.72, 2.35, 3.08, 3.94),
  legal = c("SW-14-8-2E1", "NE-22-08-1E", "NW-10-9-1E1", "RL-22-St. Norbert"))
soil_soc
```

### Download the data set
Your first step is to load the package and then download the Manitoba Original Survey Legal Descriptions data from [DataMB](https://geoportal.gov.mb.ca/) which `mbquartR` helps you do.


Load `mbquartR`
```{r setup}
library(mbquartR)
```

Download data set
```{r eval = FALSE}
quarters_dl()
```

### Search

We can now use `search_legal()` to find the center coordinates (latitude and longitude) for each of the parcels of land in our data frame. To use this function you need to supply a vector that contains the legal land descriptions of interest. This will return data frame that includes the type of land division and the coordinates of the centre. 

```{r}
location_1 <- search_legal(x = soil_soc$legal)
location_1
```


### Map

We can also use `map_quarter()` to quickly plot the center coordinates on a map. Because quarter sections are a square of known dimensions a polygon is also displayed showing the approximate borders. This is not done for other land divisions because the dimensions and orientation are inconsistent.

```{r}
map_quarter(location_1)
```



## Example 2: Using coordinates to find parcels of land

As an agri-environmental scientist I take a lot of soil samples on farms as part of research projects and I always geo-locate sampling sites so I can revisit them at a later date. When reporting results back to producers/farmers it is often easiest to refer to locations by their legal land descriptions because these descriptions are familiar and are tied to how farmers manage their land. The example below presents a small data frame of surface soil sample P concentrations (ppm), all collected from the watershed, along with their corresponding coordinates.


```{r}
soil_p <- data.frame(
  lat = c(49.333854, 49.338577, 49.356048, 49.372180),
  long = c(-98.376486, -98.376862, -98.304846, -98.254691),
  p_conc = c(8.75, 12.30, 17.45, 22.60),
  sample_id = factor(seq(1, 4, 1)))
soil_p

```

### Search

Using `search_coord()` we provide coordinates, in decimal degrees, and it searches for the quarter section, or other land division type, based on the shortest euclidean distance to the centre of a parcel of land. Because the data being search is so big this function can take a little while to run. This function returns the legal land description, land division type, the center coordinates, user coordinates, and the distance between the user provided coordinates and the centre coordinates. In the example below I have collected soil samples from two different farm fields that happen to be within the same quarter section

```{r}
location_2 <- search_coord(lat = soil_p$lat, long = soil_p$long)
location_2
```

### Map

For a quick reference I can quickly plot the centre coordinates and the approximate boundaries of these quarter sections using `map_quarter()`.


```{r}
map_quarter(location_2)
```

I can also add my sampling locations to the map by first converting my original data frame to an `sf` object.
```{r}
point <- sf::st_as_sf(soil_p, coords = c("long", "lat"),
                      crs = "+proj=longlat +datum=WGS84")
```

Then adding this `sf` object as an additional layer

```{r}
map_quarter(location_2) + 
  mapview::mapview(point, col.regions = "black", 
                   color = "black", legend =FALSE)
```



```{r, include=FALSE}
options("mbquartR_example" = FALSE)
```
